---
output:
  md_document:
    variant: markdown_github
references:
- id: Touloumis2015
  title: "R Package multgee: A Generalized Estimating Equations Solver for Multinomial Responses"
  author:
  - family: Touloumis
    given: Anestis
  container-title: Journal of Statistical Software
  volume: 64
  URL: 'https://www.jstatsoft.org/v064/i08'
  DOI: 10.18637/jss.v064.i08
  issue: 3
  page: 1-14
  type: article-journal
  issued:
    year: 2015
- id: Touloumis2013
  title: "GEE for multinomial responses using a local odds ratios parameterization"
  author:
  - family: Touloumis
    given: Anestis
  - family: Agresti
    given: Alan
  - family: Kateri
    given: Maria
  container-title: Biometrics
  volume: 69
  URL: 'http://onlinelibrary.wiley.com/doi/10.1111/biom.12054/full'
  DOI: 10.1111/biom.12054
  issue: 3
  page: 633--640
  type: article-journal
  issued:
    year: 2013
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
library("badger")
knitr::opts_chunk$set(
  tidy= TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# multgee: GEE Solver for Correlated Nominal or Ordinal Multinomial Responses


[![Travis-CI Build Status](https://travis-ci.org/AnestisTouloumis/multgee.svg?branch=master)](https://travis-ci.org/AnestisTouloumis/multgee)
`r badge_devel("AnestisTouloumis/multgee", "brightgreen")`
[![Project Status: Active The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) 
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)


[![CRAN Version](http://www.r-pkg.org/badges/version/multgee?color=blue)](https://cran.r-project.org/package=multgee)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/multgee?color=blue)](http://cranlogs.r-pkg.org/badges/grand-total/multgee)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/multgee)](http://cran.rstudio.com/web/packages/multgee/index.html)


## Author

Anestis Touloumis <https://brighton.academia.edu/AnestisTouloumis>

School of Computing, Engineering and Mathematics, University of Brighton. 


## Installation

You can install the release version of the **multgee** package from CRAN:

```{r eval=FALSE}
install.packages("multgee")
```

or the development version from github:

```{r eval=FALSE}
install.packages("devtools") # if you have not installed "devtools" package
devtools::install_github("AnestisTouloumis/multgee")
```

The source code for the **multgee** package is available on github at

- https://github.com/AnestisTouloumis/multgee.

To use **multgee**, you should load the package as follows:

```{r}
library(multgee)
```

## Usage and functions

The **multgee** package provides a generalized estimating equations (GEE) solver for fitting marginal regression models with correlated nominal or ordinal multinomial responses based on a local odds ratios parameterization for the association structure [@Touloumis2015].


The package contains two core functions that fit GEE models for correlated multinomial responses:

- `ordLORgee` for an ordinal response scale. Options for the marginal model include cumulative link models or an adjacent categories logit model,
- `nomLORgee` for a nominal response scale. A marginal baseline category logit model can be fitted.

The main arguments in both functions are: 

- an optional data frame `data`,
- a model formula `formula`,
- a cluster identifier variable `id`,
- an optional vector that identifies the order of the observations within each cluster `repeated`.


The association structure among the correlated multinomial responses is expressed via marginalized local odds ratios [@Touloumis2013]. The estimating procedure for the local odds ratios can be summarized as follows: For each level pair of the `repeated` variable, the available responses are aggregated across clusters to form a square marginalized contingency table. Treating these tables as independent, an RC-G(1) type model is fitted in order to estimate the marginalized local odds ratios. The `LORstr` argument determines the form of the marginalized local odds ratios structure. Since the general RC-G(1) model is closely related to the family of association models, one can instead fit anassociation model to each of the marginalized contingency tables by setting `LORem="2way"` in the core functions. 


The **multgee** package offers two utility functions:

- `waldts` for assessing the goodness-of-fit of two nested GEE models based on a Wald test statistic,
- `intrinsic.pars` for assessing whether the underlying association structure does not change dramatically across the level pairs of `repeated`.


## Example

The following R code illustrates how to use the core functions of **multgee**.

```{r}
data(arthritis)
fitord <- ordLORgee(y~factor(time)+factor(trt)+factor(baseline), data=arthritis,
          id=id, repeated=time)
summary(fitord) 
```

## Getting help
Details about the methods implemented in *multgee* can be found in @Touloumis2013. A detailed description of *multgee* can be found in @Touloumis2015. An updated version of this paper also serves as vignette for *multgee* which can be accessed by executing the following R commands: 

```{r, eval=FALSE}
browseVignettes("multgee")
```


## How to cite

```{r}
citation("multgee")
```

# References